Author disambiguation of large-scale semi-labeled collections of scientific articles

### 2015-08-14 paper.tex
blocking and implementation
Signed-off-by: Mateusz Susik <mateusz.susik@cern.ch>

% For algorithms
\usepackage{algorithm}
\usepackage{algorithmic}

@@

@@

\glnote{By Mateusz}

% motivation
% last name, first initial
% phonetics something (dm)

\subsection{Learning a similarity function}

@@

\glnote{By Mateusz}

% cite sklearn paper
% cite sklearn api paper

@@

\subsection{Blocking}
% compare with torvik & smalheiser (2009) who argue that lnfi blocking has recall around 98%

>>>>

% For algorithms
\usepackage{algorithm}
\usepackage{algorithmicx}
\usepackage[noend]{algpseudocode}

@@

% For "Jabłoński"
\usepackage[utf8]{inputenc}

% For placing tables in correct place
\usepackage{placeins}

@@

If the goal of the algorithm is to disambiguate millions of signatures and the algorithm
complexity is $\Omega{(n^2)}$ in terms of the number of signatures due to pairwise
comparisons, the disambiguation might take weeks or months to finish. In order to prevent
that, we \textit{block} the signatures \citep{fellegi69}. The simplest strategy groups the
signatures together if they share the same surnames and first given name initial (e.g.,
"Wang, Z"). As the clustering will be performed only within the
blocks, such strategy might prevent the further steps of the algorithm from assigning
correctly two signatures to the same author. We extracted from our training set all pairs
where the signatures belong to different blocks and the same author after the last name and
first initial strategy was used. We observed that we can classify majority of these pairs
into few categories:
\begin{enumerate}
  \item Author has multiple surnames and on some of his signatures all the surnames but the
  last one are contained within given names (e.g., "Martinez Torres, A." and "Torres, A.
  Martinez").
  \item Author has multiple surnames and on some of his signatures only the first surname is
  present (e.g., "Smith-Jones, A." and "Smith, A.")
  \item Author comes from a country that doesn't use latin alphabet and there are different
  ways of transcripting his name or his name sometimes contains a typo on the signatures
  (e.g., "Mueller, R." and "Muller, R.", "Tchaikovsky, P." and "Czajkowski, P.").
  \item Author sometimes uses on the signature a different name than his first one (e.g.,
  "Smith, Jack" and "Smith, A.J.")
  \item Author changed his surname (e.g., because of marrying somebody).
\end{enumerate}

We tackled first four categories\msnote{Should I mention why not all of them?} and
implemented a blocking strategy that produces less pairs with signatures from the same
author but different blocks, and the blocks output by the algorithm are small. Before
performing blocking, the names on the signatures are \textit{normalized}: the accents are
removed (e.g., "Jabłoński, Ł" $\rightarrow$ "Jablonski, L") and common affixes are removed
(e.g., "van der Waals, J. D."  $\rightarrow$ "Waals, J. D.")\msnote{Better explanation of
"affix" needed?}.
\begin{algorithm}
  \caption{Blocking using phonetic algorithm}
  \begin{algorithmic}[1]
    \newcommand{\myindent}[1]{
        \newline\makebox[#1cm]{}
    }
    \Require{$S$ - set of signatures. The names on the signatures should be
             \textit{normalized},
             \myindent{0}$phonetic\_surnames$ - functions that outputs phonetic
             representation of surnames from \myindent{1.38}given signature,
             \myindent{0}$given\_names$ - function that ouputs all the given names from
             given signature.}

    \Function{Phonetic blocking}{$S$}
      \State $blocks\gets \emptyset$
      \Comment{Blocks are indexed by a phonetic representation of a surname}
      \ForAll{$s$ in $S$ with single surname}
         \myindent{0.92}assign $s$ to the block of $phonetic\_surnames(s)[0]$
      \EndFor
      \ForAll{$s$ in $S$ with multiple surnames}
         \If{block of $phonetic\_surnames(s)[-1]$ contains a signature $t$
         \myindent{1.84}where $phonetic\_surnames(s)[0] = given\_names(t)[-1]$}
         \myindent{1.38}assign $s$ to the block of $phonetc\_surnames(s)[-1]$
         \ElsIf{block of $phonetic\_surnames(s)[0]$ exists}
          \myindent{1.38}assign $s$ to the block of $phonetic\_surnames(s)[0]$
         \Else
          \myindent{1.38}assign $s$ to the block of $phonetic\_surnames(s)[-1]$
         \EndIf
      \EndFor
      \Return $blocks$
    \EndFunction
  \end{algorithmic}
\end{algorithm}

\msnote{Should I explain that if the block does not exist, an empty one is created?}

After this algorithm is finished, all the blocks from the result of the function above can
be additionally split over the first given name initial.

\subsection{Learning a similarity function}
\glnote{By Gilles}

@@

\glnote{By Mateusz}
\msnote{If the matching is not described in this paper, change appearances of "four elements"}

Having the algorithm split into four elements, we implemented each of them separately.
Learning a similarity function and semi-supervised clustering steps required sophisticated
machine learning algorithms which implementations transcended the scope of this project.
Thus, we used ready-to-go \textit{Scikit-learn} library \citep{scikit-learn} and
semi-supervised clustering from \textit{SciPy} \citep{scipy}. In order to keep the code
maintainable and abstract, we followed practices introduces by \textit{Scikit-learn API}
\citep{scikit-learn-API}.

Performance-wise, the clustering step appeared to be the bottleneck. We improved the
algorithmic efficiency by parallelizing it. Having $n$ cores available, one can run the
semi-supervised clustering for $n$ different blocks in the same moment. In the production
systems the disambiguation often need not be run on the whole dataset. Once the dataset has
been disambiguated once, it is possible to avoid rerunning the whole algorithm when the new
signatures arrive in the digital library. First of all, the similarity function can be
reused. Moreover, the clustering has to be applied only to the blocks the new signatures are
contained within.

\msnote{Fill the table with times}
\msnote{Add times for clustering with phonetic blockings if they prove to be better}

\FloatBarrier
\begin{table}[H]
\caption{Time performance (2 GHz per CPU core, 32GB RAM)}
\centering
\begin{tabular}{|l|l|}
  \hline
  \textbf{algorithm} & \textbf{time}\\
  \hline
  blocking 1200000 signatures & dfsf \\
  \hline
  learning similarity function on 1000 pairs on reduced set of features & w1k2\\
  \hline
  learning similarity function on 1000000 pairs on reduced set of features & w2k2 \\
  \hline
  learning similarity function on 1000000 pairs on full set of features & sdfsdf \\
  \hline
  clustering 1200000 signatures on reduced set of features using 16 cores & asdsda \\
  \hline
  clustering 1200000 signatures on reduced set of features using 1 cores & aasdfs \\
  \hline
  clustering 1200000 signatures on full set of features using 16 cores & ssesdf \\
  \hline
  matching clusters from 1200000 signatures with the old onews & asdsd \\
  \hline
\end{tabular}
\end{table}

@@

\subsection{Base experiment}

The starting point of our analysis is an algorithm that blocks the signatures using the last
name and first initial strategy, samples the pairs in balanced way, uses gradient boosting
with all features included for learning the similarity function and clusters using
semi-supervised hierarchical approach.

\FloatBarrier
\begin{table}[H]
\caption{Base experiment evaluation)}
\centering
\begin{tabular}{|r|r|l|l|l|}
  \hline
  Evaluation & Set & Precision & Recall & F-score \\
  \hline
  Pairwise & test & 0.9948 & 0.9738 & 0.9842 \\
  \hline
  Pairwise & train & 0.9976 & 0.9747 & 0.986 \\
  \hline
  Pairwise & overall & 0.9952 & 0.9739 & 0.9844 \\
  \hline
  $B^3$ & test & 0.9901 & 0.976 & 0.983 \\
  \hline
  $B^3$ & train & 0.997 & 0.979 & 0.9879 \\
  \hline
  $B^3$ & overall & 0.9907 & 0.9761 & 0.9834 \\
  \hline
\end{tabular}
\end{table}

Let us introduce a term \textit{blocking limited maximum recall} to describe the highest
recall that the algorithm can theoretically reach after blocking is done.
Formally, for the set of signatures ${\cal S}$, blocking function $F: \mathcal{S}
\rightarrow \mathcal{B}$, where $ \mathcal{B} = \{s_{0}, \ldots{} , s_{N-1}\}$ is a
partition of $\mathcal{S} $ , the recall function $\mathcal{R}: \mathcal{C}_{gold},
\mathcal{C}_{predicted} \rightarrow \langle0,1\rangle$ where $\mathcal{C}_{gold}$ is the
partition of $\mathcal{S}$ where for every author there is a set in $\mathcal{C}_{gold}$
that contains only all the signatures of this author, $\mathcal{C}_{predicted}$ is the
partition of $\mathcal{S}$ output by the algorithm, and $\mathcal{Q} = \{\cup_{i=0}^{N-1}
q_i $ : $\forall_{i \in \langle0,N-1\rangle} q_i$ is a partition of $s_i$\}. Then
\textit{blocking limited maximum recall} is defined as $\max_{q \in \mathcal{Q}}
\mathcal{R}(\mathcal{C}_{gold}, q)$. For sanely defined recall functions, such as $B^{3}$
recall or \textit{pairwise} recall \msnote{Further explanation needed, why it is
equivalent?}, it is equivalent to $\mathcal{R}(\mathcal{C}_{gold}, \mathcal{B})$.

The \textit{blocking limited maximum recall} for the last name and first initial blocking
and $B^3$ scoring equals 0.9827. Comparing it to the $B^3$ recall of the base experiment,
one can see that a huge part of errors comes from the blocking step which forbids some of
the pairs to be merged together. To prevent that, we tried the phonetic blocking algorithm
using three different phonetic algorithms: Double Metaphone \citep{doublemetaphone}, NYSIIS
\citep{nysiis} and Soundex \citep{Soundex}. As Double Metaphone can return two results for a
string, we always discarded the second one when we used this algorithm.

\FloatBarrier
\begin{table}[H]
\caption{Blocking limited maximum recalls (BLMR) on the whole dataset without splitting over
the first given name initial}
\centering
\begin{tabular}{|r|l|l|}
  \hline
  blocking function & BLMR with $B^{3}$ recall & BLMR with pairwise recall \\
  \hline
  last name and first initial & 0.9827 & 0.9776 \\
  \hline
  phonetic using double metaphone & 0.9967 & 0.9953 \\
  \hline
  phonetic using nysiis & 0.9961 & 0.9951 \\
  \hline
  phonetic using soundex & 0.9966 & 0.9953 \\
  \hline
\end{tabular}
\end{table}

\msnote{Mention two new features for this experiments, other than that everything is the
same as in the base experiment}

\FloatBarrier
\begin{table}[H]
\caption{Phonetic algorithm performance}
\centering
\begin{tabular}{|r|r|r|l|l|l|}
  \hline
  Phonetic Algorithm & Split over initial & Evaluation & Precision & Recall & F-score \\
  \hline
  Double Metaphone & yes & pairwise & 0.9934 & 0.982 & 0.9876 \\
  \hline
  Double Metaphone & yes & $B^3$ & 0.9867 & 0.983 & 0.9848 \\
  \hline
  Double Metaphone & no & pairwise & - & - & - \\
  \hline
  Double Metaphone & no & $B^3$ & - & - & - \\
  \hline
  NYSIIS & yes & pairwise & 0.9942 & 0.9818 & 0.9879 \\
  \hline
  NYSIIS & yes & $B^3$ & 0.9884 & 0.9828 & 0.9856 \\
  \hline
  NYSIIS & no & pairwise & - & - & - \\
  \hline
  NYSIIS & no & $B^3$ & - & - & - \\
  \hline
  Soundex & yes & pairwise & - & - & - \\
  \hline
  Soundex & yes & $B^3$ & - & - & - \\
  \hline
  Soundex & no & pairwise & - & - & - \\
  \hline
  Soundex & no & $B^3$ & - & - & - \\
  \hline
\end{tabular}
\end{table}

\msnote{Fill the table}

% compare with torvik & smalheiser (2009) who argue that lnfi blocking has recall around 98%
% MS - recall for lnfi (blocking only) 0.9776, lnfi (whole algorithm) 0.9739,
% MS - all three results use pairwise scoring
